\htmlhr
\chapter{Annotating libraries\label{annotating-libraries}}

The Checker Framework distribution contains annotated versions of popular
libraries in directory \<checker-framework/checker/lib/>.  These currently
include
  \href{https://commons.apache.org/proper/commons-bcel/}{Apache Commons BCEL}
  and
  \href{https://github.com/arenn/java-getopt}{GNU getopt Java port}.
During type-checking you should put the annotated version of the library on
your classpath.  (This is done automatically for the JDK; you don't have to
do anything special for it.)
When running your code, you can use either the annotated
library or the regular distributed version of the library --- they
behave identically at run time.
Some projects, such as
  \href{http://plse.cs.washington.edu/daikon/}{Daikon}
  and
  \href{http://mernst.github.io/plume-lib/}{plume-lib},
are already distributed in annotated form by their maintainers;
over time, this should become more common, reducing the need for the
Checker Framework to include annotated versions of them.

%% Extra detail about the 
% \section{Using distributed annotated JDKs\label{skeleton-using}\label{skeleton}}
% 
% The Checker Framework distribution contains two
% annotated JDKs at the paths \<checker/dist/jdk7.jar> and
% \<checker/dist/jdk8.jar>.
% The \<javac> that is distributed with the Checker Framework and the command
% \code{java -jar \$CHECKERFRAMEWORK/checker/dist/checker.jar} both use the appropriate \<jdk7.jar> or \<jdk8.jar>
% based on the version of Java used to run them.
% 
% The annotated JDKs should \emph{not} be in your classpath at run time, only
% at compile time; and they are automatically added to your classpath at
% compile time.


If your code uses a library that does \emph{not} contain type annotations,
then the type-checker has no way to know the library's behavior.
The type-checker
makes conservative assumptions about unannotated bytecode:  it assumes that
every method parameter has the bottom type annotation and that every method
return type has the top type annotation (see
Section~\ref{defaults-classfile} for details and an example).
These conservative library
annotations invariably lead to checker warnings.
This chapter describes how to eliminate
the warnings by adding annotations to the library.
(Alternately, you can instead
suppress all warnings related to an unannotated library by use of the
\code{-AskipUses} or \code{-AonlyUses} command-line option; see
Section~\ref{askipuses}.)

(Note: This chapter uses ``library'' to refer to code that is provided in
\<.class> or \<.jar> form.  You should use this approach for parts of your
own codebase if you typically compile different parts separately.  When
your codebase is typically compiled together and you are type-checking only
part of it, you can use the approach described in this chapter, or you can
use command-line arguments such as \<-AskipUses> and \<-AskipDefs> (see
Sections~\ref{askipuses}--\ref{askipdefs}).  Also, recall that the Checker
Framework analyzes all, and only, the source code that is passed to it.
The Checker Framework is a plug-in to the javac compiler, and it never
analyzes code that is not being compiled, though it does look up
annotations in the class files for code that is not being compiled.)

You make the library's annotations known to the checkers by writing
annotations in a copy of the library's source code.
If you do not have access to the source code, you can create and use a ``stub
file'' instead.
Given the library annotations, you have two options:

\begin{enumerate}
\item
  You can compile the library to create \<.class> and \<.jar> files that
  contain the annotations.  Then, when doing pluggable type-checking, you
  would put those files on the classpath.
  When running your code, you can use either version of the library:  the
  one you created or the original distributed version.

  %% There is no point to advertising this deprecated workflow.
  % You can insert annotations in the compiled \code{.class} files of the
  % library.  You would express the annotations textually, typically as an
  % annotation index file, and
  % then insert them in the library by using the Annotation File Utilities
  % (\myurl{http://types.cs.washington.edu/annotation-file-utilities/}).
  % See the Annotation File Utilities documentation for full details.

  With this compilation approach, the syntax of the library annotations is
  validated ahead of time.  This makes the compilation approach less
  error-prone, and it makes the type-checker run faster.  It gives
  correctness guarantees about the library in addition to your code.
  Section~\ref{compiling-libraries} describes how to compile a library.

\item
  You can supply the annotated library source code, or a very concise
  variant called a ``stub file'', textually to the Checker Framework.

  The stub file approach does not require you to compile the library source
  code.
  A stub file is applicable to multiple versions of a library, so
  the stub file does not need to to be updated when a new version of the
  library is released.
  When provided by the author of the checker, a stub file is used
  automatically, with no need for the user to supply a command-line option.
  The stub file reader approach has some limitations, notably using
  non-standard syntax in some locations (Section~\ref{stub-limitations}).
  Section~\ref{stub} describes how to create and use stub files.

\end{enumerate}


% It uses both of the above mechanisms.  The
% Nullness, Javari, IGJ, and Interning Checkers use the annotated JDKs
% (Section~\ref{skeleton}), and all other checkers use stub files
% (Section~\ref{stub}).

If you write any library annotations,
please share them so that they
can be distributed with the Checker Framework.
Sharing your annotations is useful even if the library is only partially
annotated.
%% This text does not feel so relevant.
% You can determine the correct annotations for a library either
% automatically by running an inference tool, or manually by reading the
% documentation.  Presently, automated type inference tools are available for the
% Nullness (Section~\ref{nullness-inference}) and Javari
% (Section~\ref{javari-inference}) type systems.


\section{Compiling partially-annotated libraries\label{compiling-libraries}}

If you completely annotate a library, then you can compile it using a
pluggable type-checker, and include the resulting \<.jar> file on your
classpath.  You get a guarantee that the library contains no errors.

The rest of this section tells you how to compile a library if you
\emph{partially} annotate it, such as writing annotations for some of its
classes but not others.
When compiling a partially-annotated library, the checker needs to use normal
defaulting rules (Section~\ref{climb-to-top}) for code you have annotated and
conservative defaulting rules (Section~\ref{defaults-classfile}) for
code you have not yet annotated.
You use \<@AnnotatedFor> to indicate which classes you have annotated.

When compiling a library that is not fully annotated, use command-line
argument \<-AuseConservativeDefaultsForUnannotatedCode>.  This causes the
checker to assume that all source files being compiled are unannotated
libraries, except any classes with a relevant \<@AnnotatedFor> annotation.
In particular, the checker uses conservative defaults (see
Section~\ref{defaults-classfile}) for any type use with no explicit
user-written annotation, and the checker issues no warnings.

The \refqualclass{framework/qual}{AnnotatedFor} annotation, written on a
class, indicates that the class has been annotated for certain type
systems.  For example, \<@AnnotatedFor(\ttlcb"nullness", "regex"\ttrcb)> means that
the programmer has written annotations for the Nullness and Regular
Expression type systems.  If one of those two type-checkers is run, this
class is treated normally, even if the
\<-AuseConservativeDefaultsForUnannotatedCode> command-line argument is
supplied:  unannotated types are defaulted using normal source-code
defaults and type-checking warnings are issued.
\refqualclass{framework/qual}{AnnotatedFor}'s arguments are any string that
may be passed to the \<-processor> command-line argument:  the
fully-qualified class name for the checker, or a shorthand for built-in
checkers (see Section~\ref{shorthand-for-checkers}).

Whenever you compile a class using the Checker Framework, including when
using the \<-AuseConservativeDefaultsForUnannotatedCode> command-line
argument, the resulting \<.class> files are fully-annotated; each type use
in the \<.class> file has an explicit type qualifier for any checker that
is run.


\subsection{Workflow for creating or augmenting a partially-annotated library\label{compiling-libraries-workflow}}

This section describes the typical workflow for creating a
partially-annotated library.

\begin{enumerate}
\item Find out whether an annotated version of the library already exists
  (see file \<checker-framework/checker/lib/README>).
%% This adds clutter to the source code, so omit the step from these instructions.
% \item Optionally, run a script that adds
%     \<\refqualclass{framework/qual}{AnnotatedFor}(\ttlcb\ttrcb)>
%     to each class.  [[TODO: say where to find this script.]]
%     This step has no semantic effect.  It only saves you the trouble of
%     typing those 17 characters later.

  If it does not already exist, fork the project (if its license permits forking).
  Adjust the library's
  build process, such as a Maven or Ant buildfile, so that when running the
  compiler, it passes the \<-AuseConservativeDefaultsForUnannotatedCode>
  command-line option and runs every pluggable type-checker for which any
  annotations exist.  The reason to run every type-checker is so that all
  of the annotations are defaulted properly.

\item Annotate some files.
  You can determine which files need to be annotated by using the
  \<-AprintUnannotatedMethods> command-line argument while type-checking a
  client of the library.

  % This is a strong recomendation, but not a requirement.  @AnnotatedFor
  % can be written on a method as well.
  When you annotate a file, annotate the whole thing, not just a few of its
  methods.  Once the file is annotated, add an
  \<\refqualclass{framework/qual}{AnnotatedFor}(\ttlcb"\emph{checkername}"\ttrcb)>
  annotation to its classes, or augment an existing \<@AnnotatedFor>
  annotation.

\item
  Build the library.
  Now you have a \<.jar> file that you can use while type-checking and at
  run time.

\item
  Tell other people about your work so that they can benefit from it.

  \begin{itemize}
  \item
    Please inform the Checker Framework developers
    about your new annotated library by opening an issue or issuing a pull
    request.  This will let us include your annotated library in directory
    \<checker-framework/checker/lib> of the Checker Framework release.

  \item
    Encourage the library's maintainers to accept your annotations into its
    main version control repository.  This will make the annotations easier
    to maintain, the library will obtain the correctness guarantees of
    pluggable type-checking, and there will be no need for the Checker
    Framework to include an annotated version of the library.

    You will probably want to write the annotations in comments, so that it
    is still possible to compile the library without use of Java 8.

    If the library maintainers do not accept the annotations, then
    periodically, such as when a new version of the library is released,
    pull changes from upstream (the library's main version control system)
    into your fork, add annotations to any newly-added methods in classes
    that are annotated with \<@AnnotatedFor>, rebuild to create an updated
    \<.jar> file, and inform the Checker Framework developers by opening an
    issue or issuing a pull request.
  \end{itemize}

\end{enumerate}


\section{Using stub classes\label{stub}\label{stub-creating-and-using}}

A stub file contains ``stub classes'' that contain annotated signatures,
but no method bodies.  A
checker uses the annotated signatures at compile time, instead of or in
addition to annotations that appear in the library.

Section~\ref{stub-creating} describes how to create stub classes.
Section~\ref{stub-using} describes how to use stub classes.
These sections illustrate stub classes via the example of creating a \refqualclass{checker/interning/qual}{Interned}-annotated
version of \code{java.lang.String}.  You don't need to repeat these steps
to handle \code{java.lang.String} for the Interning Checker,
but you might do something similar for a different class and/or checker.

% First, you must install the skeleton class generator
% (Section~\ref{skeleton-installing}).


\subsection{Using a stub file\label{stub-using}}

The \code{-Astubs} argument causes the Checker Framework to read
annotations from annotated stub classes in preference to the unannotated
original library classes.  For example:

%BEGIN LATEX
\begin{smaller}
%END LATEX
\begin{Verbatim}
  javac -processor org.checkerframework.checker.interning.InterningChecker -Astubs=String.astub:stubs MyFile.java MyOtherFile.java ...
\end{Verbatim}
%BEGIN LATEX
\end{smaller}
%END LATEX

Each stub path entry is a file or a directory; specifying a directory is
equivalent to specifying every file in it whose name ends with
\code{.astub}.  The stub path entries are delimited by
\<File.pathSeparator> (`\<:>' for Linux and Mac, `\<;>' for Windows).

A checker automatically reads the stub file \code{jdk.astub}, unless
command-line option \<-Aignorejdkastub> is supplied.  (The checker
author should place \code{jdk.astub} in the same directory as the Checker class, i.e.,
the subclass of \code{BaseTypeVisitor}.)  Programmers should only use the
\<-Astubs> argument for additional stub files they create themselves.

If a method appears in more than one stub file (or twice in the same 
stub file), then the annotations are merged. If any of the 
methods have different annotations from the same hierarchy on the same type,
then the annotation from the last declaration is used.  

% \textbf{The following is not implemented yet}
% A library writers should create a file \code{library.astub} on the
% classpath (in the resources directory or the binary jars).
% The Checker Framework automatically imports all the stub files named
% \code{library.astub} found in the classpath.  



\subsection{Stub file format\label{stub-format}}

Every Java file is a valid stub file.  However, you can omit information
that is not relevant to pluggable type-checking; this makes the stub file
smaller and easier for people to read and write.

As an illustration, a stub file for the Interning type system
(Chapter~\ref{interning-checker}) could be:

\begin{Verbatim}
  import org.checkerframework.checker.interning.qual.Interned;
  package java.lang;
  @Interned class Class<T> { }
  class String {
    @Interned String intern();
  }
\end{Verbatim}

Note, annotations in comments are ignored.

The stub file format is allowed to differ from Java source code in the
following ways:
\begin{description}

\item{\textbf{Method bodies:}}
  The stub class does not require method bodies for classes; any method
  body may be replaced by a semicolon (\code{;}), as in an interface or
  abstract method declaration.

\item{\textbf{Method declarations:}}
  You only have to specify the methods that you need to annotate.
  Any method declaration may be omitted, in which case the checker reads
  its annotations from library's \<.class> files.  (If you are using a stub class, then
  typically the library is unannotated.)

\item{\textbf{Declaration specifiers:}}
  Declaration specifiers (e.g., \<public>, \<final>, \<volatile>)
  may be omitted.

\item{\textbf{Return types:}}
  The return type of a method does not need to match the real method.
  In particular, it is valid to use \<java.lang.Object> for every method.
  This simplifies the creation of stub files.

\item{\textbf{Import statements:}}
  All imports must be at the beginning of the file.
  The only required import statements are the ones to import type
  annotations.  Import statements for types are optional.

  Enum constants in annotations need to be either fully qualified
  or imported.
  For example, one has to either write the enum constant \<ANY> in
  fully-qualified form:

\begin{Verbatim}
@Source(sparta.checkers.quals.FlowPermission.ANY)
\end{Verbatim}

\noindent
or correctly import the enum class:

\begin{Verbatim}
import sparta.checkers.quals.FlowPermission;
...
@Source(FlowPermission.ANY)
\end{Verbatim}

\noindent
or statically import the enum constants:

\begin{Verbatim}
import static sparta.checkers.quals.FlowPermission.*;
...
@Source(ANY)
\end{Verbatim}

  Importing all packages from a class (\<import my.package.*;>) only
  considers annotations from that package; enum types need to be
  explicitly imported.

\item{\textbf{Multiple classes and packages:}}
  The stub file format permits having multiple classes and packages.
  The packages are separated by a package statement:
  \<package my.package;>.  Each package declaration may occur only once; in
  other words, all classes from a package must appear together.

\end{description}



\subsection{Creating a stub file\label{stub-creating}}


\subsubsection{If you have access to the Java source code\label{stub-creating-with-source}}

Every Java file is a stub file.  If you have access to the Java file, then
you can use the Java file as the stub file.  Just add
annotations to the signatures, leaving the method bodies unchanged.
The stub file parser silently ignores any annotations that it cannot
resolve to a type, so don't forget the \<import> statement.

Optionally (but highly recommended!), run the type-checker to verify that
your annotations are correct.  When you run the type-checker on your
annotations, there should not be any stub file that also contains
annotations for the class.  In particular, if you are type-checking the JDK
itself, then you should use the \<-Aignorejdkastub> command-line option.

This approach retains the original
documentation and source code, making it easier for a programmer to
double-check the annotations.  It also enables creation of diffs, easing
the process of upgrading when a library adds new methods.  And, the
annotations are in a format that the library maintainers can even
incorporate.

The downside of this approach is that the stub files are larger.  This can
slow down parsing.
% Furthermore, a programmer must search the stub file
% for a given method rather than just skimming a few pages of method signatures.


\subsubsection{If you do not have access to the Java source code\label{stub-creating-without-source}}

If you do not have access to the library source code, then you can create a
stub file from the class file (Section~\ref{stub-creating}),
and then annotate it.  The rest of this section describes this approach.


\begin{enumerate}

\item
  Create a stub file by running the stub class generator.  (\<checker.jar> and \<javac.jar>
  must be on your classpath.)

\begin{Verbatim}
  cd nullness-stub
  java org.checkerframework.framework.stub.StubGenerator java.lang.String > String.astub
\end{Verbatim}

  Supply it with the fully-qualified name of the class for which you wish to
  generate a stub class.  The stub class generator prints the
  stub class to standard out, so you may wish to redirect its output to a
  file.

\item
  Add import statements for the annotations.  So you would need to
add the following import statement at the beginning of the file:

\begin{Verbatim}
  import org.checkerframework.checker.interning.qual.*;
\end{Verbatim}

\noindent
The stub file parser silently ignores any annotations that it cannot
resolve to a type, so don't forget the import statement.
Use the \<-AstubWarnIfNotFound> command-line option to see warnings
if an entry could not be found.

\item
  Add annotations to the stub class.  For example, you might annotate
  the \sunjavadoc{java/lang/String.html\#intern--}{String.intern()} method as follows:

\begin{Verbatim}
  @Interned String intern();
\end{Verbatim}

  You may also remove irrelevant parts of the stub file; see
  Section~\ref{stub-format}.

\end{enumerate}


\subsection{Troubleshooting stub libraries\label{stub-troubleshooting}}


\subsubsection{Type-checking does not yield the expected results}

By default, the stub parser silently ignores
annotations on unknown classes and methods.
The stub parser also silently ignores unknown annotations, so don't forget to
\<import> any annotations.  

Use command-line option 
\<-AstubWarnIfNotFound> to warn whenever some element of a stub file cannot
be found.

The \<@NoStubParserWarning> annotation on a package or type in a stub file
overrides the \<-AstubWarnIfNotFound> command-line option, and no warning
will be issued.

Use command-line option \<-AstubDebug> to output debugging messages while
parsing stub files, including about unknown classes, methods, and
annotations.  This overrides the \<@NoStubParserWarning> annotation.



\subsubsection{Problems parsing stub libraries}

When using command-line option \<-AstubWarnIfNotFound>,
an error is issued if a stub file has a typo or the API method does not
exist.

Fix this error by removing the extra L in the method name:
\begin{Verbatim}
StubParser: Method isLLowerCase(char) not found in type java.lang.Character
\end{Verbatim}

Fix this error by removing the method \<enableForgroundNdefPush(...)> from
the stub file, because it is not defined in class \<android.nfc.NfcAdapter>
in the version of the library you are using:
\begin{Verbatim}
StubParser: Method enableForegroundNdefPush(Activity,NdefPushCallback) 
      not found in type android.nfc.NfcAdapter
\end{Verbatim}


\subsection{Limitations\label{stub-limitations}}

The stub file reader has several limitations.  We will fix these in a
future release.

\begin{itemize}

\item 
  The receiver is written after the method parameter list, instead of as an
  explicit first parameter.  That is, instead of

\begin{Verbatim}
     returntype methodname(@Annotations C this, params);
\end{Verbatim}

\noindent
in a stub file one has to write

\begin{Verbatim}
     returntype methodname(params) @Annotations;
\end{Verbatim}

\item
  The stub file reader
  does not handle nested classes.  To work around this, it permits a
  top-level class to be written with a \<\$> in its name, and applies the
  annotations to the appropriate nested class.

\item
  Annotations must be written before the package name on a fully qualified 
  types rather than directly on the type it qualifies.  However, it is usually not 
  necessary to write the fully qualified name.  
  
  \begin{Verbatim}
          void init(@Nullable java.security.SecureRandom random);
   \end{Verbatim}

\item
  Annotations can only use string, boolean, or integer literals; other literals are not
  yet supported.
\end{itemize}

If these limitations are a problem, then you should insert annotations
in the library's \<.class> files instead.


% Label "skeleton" is for old links from the Javarifier manual, to prevent
% them from being broken links.


% Skeleton classes are inferior to stub classes for two reasons.  First,
% skeleton files must be on the classpath during compilation but must
% \emph{not} be on the classpath during execution; this is inconvenient and
% error-prone.  Second, the skeleton files contain incorrect values for
% certain static final fields.  These incorrect values can lead to
% run-time problems unless the Java code is re-compiled without the skeleton
% classes after type-checking is complete.



% \section{Installing the skeleton class generator\label{skeleton-installing}}
%
% Source code for the skeleton class generator tool is included in the
% Checker Framework
% distribution, but because the tool has additional dependencies, the provided
% build script does not build the tool by default.
%
% Follow these steps to install the skeleton class generator:
%
% \begin{enumerate}
%
% \item
%   Install the annotation file utilities, using the instructions at
%   \myurl{http://types.cs.washington.edu/annotation-file-utilities/}.
%   Per those instructions, the \code{annotation-file-utilities.jar} file
%   should be on your classpath.
%
% % TODO This item should become optional; tell people to install the AFU in
% % the right place.
% \item
%   Update the \code{build.properties} file in the Checker Framework distribution so
%   that the \code{annotation-utils.lib} property specifies the location of
%   the \code{annotation-file-utilities.jar} library.
%
% \item
%   Build the skeleton class generator tool by running \code{ant
%     skeleton-util dist} in the \code{checker} directory.  This updates the
%   \code{checker.jar} file to contain the skeleton class generator.
%   \code{checker.jar} should already be on your classpath (see
%   Section~\ref{installation}).
%
% \end{enumerate}


\section{Troubleshooting/debugging annotated libraries\label{libraries-troubleshooting}}

Sometimes, it may seem that a checker is treating a library as unannotated
even though the library has annotations.  The compiler has two flags that
may help you in determining whether library files are read, and if they are
read whether the library's annotations are parsed.

\begin{description}
\item \<-verbose>
  Outputs info about compile phases --- when the compiler
  reads/parses/attributes/writes any file.  Also outputs the classpath and
  sourcepath paths.
\item \<-XDTA:parser> (which is equivalent to \<-XDTA:reader> plus \<-XDTA:writer>)
  Sets the internal \<debugJSR308> flag, which outputs information about
  reading and writing.
\end{description}


% LocalWords:  plugin utils util dist RuntimeException NonNull TODO AFU enum
% LocalWords:  sourcepath Nullness javac classpath src quals pathSeparator JDKs
% LocalWords:  IGJ's jdk Astubs skipUses astub AskipUses toArray IGJ JDK6
% LocalWords:  CollectionToArrayHeuristics BaseTypeVisitor Xbootclasspath
% LocalWords:  Interning's UsesObjectEquals Anocheckjdk AonlyUses java
%  LocalWords:  Aignorejdkastub AstubWarnIfNotFound AstubDebug jdk6 jdk7
%  LocalWords:  enableForgroundNdefPush XDTA debugJSR308 BCEL getopt jdk8
%%  LocalWords:  NoStubParserWarning CHECKERFRAMEWORK AnnotatedFor regex
%%  LocalWords:  AuseConservativeDefaultsForUnannotatedCode buildfile
%%  LocalWords:  AprintUnannotatedMethods checkername AskipDefs
